---
title: "RCMEM2 Examples"
author: "James R Holmquist"
date: "2026-03-11"
output: html_document
---

## Things you can do in version 2

First, let's set up some scenarios to run some comparisons over. These are from Annapolis MD and are based on the Ecological Monographs paper in review.

```{r parameters}

  library(tidyverse)

  # SLR
  annapolis_gap_filled <- read_csv("../inst/extdata/sea_level/annapolis_gap_filled.csv")

  loess_model <- loess(msl_cm~year, data = annapolis_gap_filled)
  annapolis_gap_filled$loess_msl <- predict(loess_model, data.frame(year = annapolis_gap_filled$year))

  meanSeaLevel = annapolis_gap_filled$loess_msl
  meanSeaLevelDatum = -1.958671252
  meanHighWaterDatum = 5.521639027
  meanHighHighWaterDatum = 19.81578852
  meanHighHighWaterSpringDatum=26.25399172
  lunarNodalAmp = 0
  # suspendedSediment = 22 * 1e-06, # assume 75.8 mg/l in creek and interior is 1/3 to 1/4 creek

  meanHighWater <- meanHighWaterDatum-meanSeaLevelDatum
  meanHighHighWater <- meanHighHighWaterDatum-meanSeaLevelDatum
  meanHighHighWaterSpring <- meanHighHighWaterSpringDatum-meanSeaLevelDatum

  # suspendedSediment = 10 * 1e-06 # original
  nFloods <- 3
  flood_frequency <- c(0.5, 0.46497542, 0.03502458) * 705.79
  msl <- meanSeaLevel
  mhwMat <- matrix(c(msl+meanHighWater,msl+meanHighHighWater,msl+meanHighHighWaterSpring),
                   ncol = 3)
  mlwMat <- msl - (mhwMat-msl)

  roxygen2::roxygenise("..")
  
```

## Batching Scenarios

We can now batch multiple years into a single cohort. This allows us to modulate run time. Run time increases geometrically with each cohort added. Batching multiple years in a single cohort means lower run time.

```{r batching scenario}

run_times <- c(NA, NA, NA, NA, NA)

start <- Sys.time()

scenario_1 <- runCohortMem2(run_spinup = 1, 
                                  run_scenario = 1,
                                  msl = msl,
                                  mhwMat = mhwMat,
                                  mlwMat = mlwMat,
                                  spinup_iterations = 200, 
                                  years_per_spinup_iter = 1,
                                  rootShape = c(0,0))[[2]] %>% 
  mutate(spinup_settings = "1 year iter, 200 iterations")

end <- Sys.time()

run_times[1] <- end-start

start <- Sys.time()

scenario_2 <- runCohortMem2(run_spinup = 1, 
                                  run_scenario = 1,
                                  msl = msl,
                                  mhwMat = mhwMat,
                                  mlwMat = mlwMat,
                                  spinup_iterations = 100, 
                                  years_per_spinup_iter = 2,
                                  rootShape = c(0,0))[[2]] %>% 
  mutate(spinup_settings = "2 year iter, 100 iterations")

end <- Sys.time()

run_times[2] <- end-start

start <- Sys.time()

scenario_5 <- runCohortMem2(run_spinup = 1, 
                                  run_scenario = 1,
                                  msl = msl,
                                  mhwMat = mhwMat,
                                  mlwMat = mlwMat,
                                  spinup_iterations = 40, 
                                  years_per_spinup_iter = 5,
                                  rootShape = c(0,0))[[2]] %>% 
  mutate(spinup_settings = "5 year iter, 40 iterations")

end <- Sys.time()

run_times[3] <- end-start

start <- Sys.time()

scenario_10 <- runCohortMem2(run_spinup = 1, 
                                  run_scenario = 1,
                                  msl = msl,
                                  mhwMat = mhwMat,
                                  mlwMat = mlwMat,
                                  spinup_iterations = 20, 
                                  years_per_spinup_iter = 10,
                                  rootShape = c(0,0))[[2]] %>% 
  mutate(spinup_settings = "10 year iter, 20 iterations")

end <- Sys.time()

run_times[4] <- end-start

start <- Sys.time()


scenario_20 <- runCohortMem2(run_spinup = 1, 
                                   run_scenario = 1,
                                   msl = msl,
                                   mhwMat = mhwMat,
                                   mlwMat = mlwMat,
                                   spinup_iterations = 10, 
                                   years_per_spinup_iter = 20,
                                   rootShape = c(0,0)) [[2]] %>% 
  mutate(spinup_settings = "20 year iter, 10 iterations")

end <- Sys.time()

run_times[5] <- end-start

plot(x = c(200, 100, 40, 20, 10), y = run_times, xlab = "N initial cohorts", ylab = "Runtime",
     type = "o")

```


```{r output comparisons}

scenarios <- scenario_1 %>% 
  bind_rows(scenario_5) %>% 
  bind_rows(scenario_20) %>% 
  select(year, surface_elevation, aboveground_biomass, belowground_biomass, totalRootVolume, spinup_settings) %>% 
  pivot_longer(names_to = "variable", -c(year, spinup_settings))

ggplot(scenarios, aes(x = year, y = value)) +
  facet_wrap(.~variable, scale = "free_y") +
  geom_point(aes(color = spinup_settings, shape = spinup_settings)) + 
  scale_shape_manual(values = c(1,3,4)) +
  xlim(c(1900, 2018))

```

## We can also run multip types of competition function

Setting competition function to zero means zero-sum. If multiple sets of biological species.  

```{r plant competition functions}
# Preprogrammed
args(runCohortMem2)

species_10a <- runCohortMem2(run_spinup = 1, 
                                   run_scenario = 1,
                                   msl = msl,
                                   mhwMat = mhwMat,
                                   mlwMat = mlwMat,
                                   spinup_iterations = 20, 
                                   years_per_spinup_iter = 10,
                                   competition_function = 0,
                                   rootShape = c(1,1),
                             species_codes = c("SCAM", "SPPA"))[[3]] %>% 
  as.data.frame() %>% 
  mutate(spinup_settings = "Zero-sum")

species_10b <- runCohortMem2(run_spinup = 1, 
                                   run_scenario = 1,
                                   msl = msl,
                                   mhwMat = mhwMat,
                                   mlwMat = mlwMat,
                                   spinup_iterations = 20, 
                                   years_per_spinup_iter = 10,
                                   competition_function = 1,
                                   rootShape = c(1,1),
                              species_codes = c("SCAM", "SPPA")) [[3]] %>% 
  as.data.frame() %>% 
  mutate(spinup_settings = "Co-existance")


species_codes <- c("SCAM", "SPPA")

species_plot <- species_10b %>% 
  bind_rows(species_10a) %>% 
  gather(key = "species", value = "aboveground_biomass", -c(year, spinup_settings)) %>%
  mutate(species = factor(species, levels = rev(species_codes))) %>% 
  group_by(spinup_settings, year) %>% 
  mutate(cumulative_biomass = cumsum(aboveground_biomass))

ggplot(species_plot, aes(x = year, y = cumulative_biomass)) +
  geom_ribbon(aes(ymax = cumulative_biomass, ymin = 0, fill = species)) +
  facet_wrap(.~spinup_settings)

```

## Separate Runs and Spinups

``` {r separate runs and spinups}

mem_1 <- runCohortMem2(run_spinup = 1, 
                             run_scenario = 1,
                             msl = msl,
                             mhwMat = mhwMat,
                             mlwMat = mlwMat,
                             scenario_calendar_years = 1928:2018,
                             spinup_iterations = 20, 
                             years_per_spinup_iter = 10) [[2]] %>% 
  mutate(run = "Full run")

mem_1_cohorts <- runCohortMem2(run_spinup = 1, 
                                     run_scenario = 1,
                                     msl = msl,
                                     mhwMat = mhwMat,
                                     mlwMat = mlwMat,
                                     scenario_calendar_years = 1928:2018,
                                     spinup_iterations = 20, 
                                     years_per_spinup_iter = 10) [[1]] %>% 
  mutate(run = "Full run")

mem_2a <- runCohortMem2(run_spinup = 1, 
                              run_scenario = 0,
                              msl = msl[1],
                              mhwMat = t(mhwMat[1,]),
                              mlwMat = t(mlwMat[1,]),
                              scenario_calendar_years = 1928,
                              spinup_iterations = 20, 
                              years_per_spinup_iter = 10)

cohort_2a <- mem_2a[[1]] %>%
  filter(year == 1928)

scenario_2a <- mem_2a[[2]] %>%
  filter(year == 1928)

species_2a <- mem_2a[[3]] %>% 
  filter(year == 1928)

mem_2b <- runCohortMem2(run_spinup = 0, 
                              run_scenario = 1,
                              msl = msl,
                              mhwMat = mhwMat,
                              mlwMat = mlwMat,
                              initial_cohorts = cohort_2a,
                              initial_scenario = scenario_2a,
                              initial_species = species_2a,
)[[2]] %>% 
  mutate(run = "Cohorta and scenario run separately")

mem_2b_cohorts <- runCohortMem2(run_spinup = 0, 
                                      run_scenario = 1,
                                      msl = msl,
                                      mhwMat = mhwMat,
                                      mlwMat = mlwMat,
                                      initial_cohorts = cohort_2a,
                                      initial_scenario = scenario_2a,
                                      initial_species = species_2a,
)[[1]] %>% 
  mutate(run = "Cohorta and scenario run separately")



together <- bind_rows(mem_1,
                      mem_2b)

together_2 <- bind_rows(mem_1_cohorts,
                        mem_2b_cohorts)

ggplot(together, aes(x = year, y = surface_elevation)) +
  geom_point(aes(color = run, shape = run))+
  scale_shape_manual(values = c(1,3,4))

ggplot(together, aes(x = year, y = aboveground_biomass)) +
  geom_point(aes(color = run, shape = run))+
  scale_shape_manual(values = c(1,3,4))

```

## Can we pass a hop test?

```{r hop test, fig.width=14, fig.height=8}

mem_full_run <- runCohortMem2(run_spinup = 1, 
                             run_scenario = 1,
                             msl = msl,
                             mhwMat = mhwMat,
                             mlwMat = mlwMat,
                             scenario_calendar_years = 1928:2018,
                             spinup_iterations = 20, 
                             years_per_spinup_iter = 10,
                             competition_function = 1)



mem_2_hop <- runCohortMem2(run_spinup = 1, 
                              run_scenario = 0,
                              msl = msl[1],
                              mhwMat = t(mhwMat[1,]),
                              mlwMat = t(mlwMat[1,]),
                              scenario_calendar_years = 1928,
                              spinup_iterations = 20, 
                              years_per_spinup_iter = 10,
                             competition_function = 1)

scenario_calendar_years <- 1928:2018

for (i in 2:length(scenario_calendar_years)) {
  
  cohort_2a <- mem_2_hop[[1]] %>%
    filter(year == scenario_calendar_years[i-1])

  scenario_2a <- mem_2_hop[[2]] %>%
    filter(year == scenario_calendar_years[i-1])

  species_2a <- mem_2_hop[[3]] %>% 
    filter(year == scenario_calendar_years[i-1])

  mem_2_hop <- runCohortMem2(run_spinup = 0, 
                              run_scenario = 1,
                              msl = msl[(i-1):i],
                              mhwMat = mhwMat[(i-1):i,],
                              mlwMat = mlwMat[(i-1):i,],
                              scenario_calendar_years = scenario_calendar_years[i-1]:scenario_calendar_years[i],
                              initial_cohorts = cohort_2a,
                              initial_scenario = scenario_2a,
                              initial_species = species_2a,
                             competition_function = 1)
  
  if (i == 2) {
    
    hop_cohorts <- mem_2_hop[[1]]
    hop_scenario <- mem_2_hop[[2]]
    hop_species <- mem_2_hop[[3]]
    
  } else {
    
    hop_cohorts <- hop_cohorts %>% bind_rows(mem_2_hop[[1]] %>% filter(year == scenario_calendar_years[i]))
    hop_scenario <- hop_scenario %>% bind_rows(mem_2_hop[[2]] %>% filter(year == scenario_calendar_years[i]))
    hop_species <- hop_species %>% bind_rows(mem_2_hop[[3]] %>% filter(year == scenario_calendar_years[i]))
    
  }
  
  
  
}

mem_full_run_extended <- calculateCarbonFlux(
  cohorts = mem_full_run[[1]]  %>% filter(year >= 1928),
  scenario = mem_full_run[[2]]  %>% filter(year >= 1928)
  )[[1]] %>% 
  mutate(run = "Continuous Run") %>% 
  left_join( mem_full_run[[3]])

hop_test_extended <- calculateCarbonFlux(cohorts = hop_cohorts,
                                         scenario = hop_scenario)[[1]] %>% 
   mutate(run = "Hop Test") %>% 
  left_join(hop_species)

display_hop <- mem_full_run_extended %>% 
  bind_rows(hop_test_extended) %>% 
  pivot_longer(-c(year, run))

ggplot(display_hop, aes(x = year, y = value)) +
 # geom_point(aes(color = run, shape = run)) +
  geom_line(aes(color = run, lty = run)) +
  facet_wrap(.~name, scale = "free") +
  #scale_shape_manual(values = c(3,4)) +
  scale_linetype_manual(values = c(1,2))


```

## We can simulate radioisotope dates

``` {r simulate radioisotopes}

a_mem_run <- runCohortMem2(run_spinup = 1, 
                           run_scenario = 1,
                           msl = msl,
                           mhwMat = mhwMat,
                           mlwMat = mlwMat,
                           spinup_iterations = 200, 
                           years_per_spinup_iter = 20,
                           rootShape = c(0),
                           omDecayRateSlow = 0.0001,
                           bMax = c(0.0622),
                           rootTurnover = c(0.5),
                           initElv = -11,
                           recalcitrantFrac = 0.5,
                           rootDepthMax = c(40),
                           captureRate = 5,
                           suspendedSediment = 10 * 1e-06,
                           species_codes = "SCAM",
                           zVegMax = 1.8,
                           zVegPeak = 1.5, 
                           zVegMin = - 1.4,
                           abovegroundTurnover = 1.5,
                           rootPackingDensity = 0.07,
                           rootToShoot = 1.7
                           )

scenario <- a_mem_run[[2]]
cohorts <- a_mem_run[[1]]

head(a_mem_run[[1]])

a_core <- buildSoilCore(cohorts = a_mem_run[[1]], coreYear = 2018, coreDepth = 200)

head(a_core)

# We need to pick a Cs 137 regional curve
args(simulateRadioisotopeDates)

# There are three curves prepackaged
listCs137Scenarios()

# Check them out
scenario_dir <- system.file(
    "extdata",
    "fallout",
    package = "rCMEM"
  )
  
files <- list.files(
  scenario_dir,
  pattern = "\\.Rds$",
  full.names = T
)
for (i in 1:length(files)) {
  
  cs_file <- readRDS(files[i])
  
  if (i == 1) {
    cs_file_comp <- cs_file
  } else {
    cs_file_comp <- bind_rows(cs_file_comp, cs_file)
  }
  
}

ggplot(cs_file_comp, aes(x = year, y = cs137_influx_Bq_m2_yr)) +
  geom_line(aes(color = Europe_or_non, lty=N_or_S))
 
a_dates <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 0.01, 
                                     cs137_mobility_factor = 0.01,
                                     cs137_erosion_loss = 0.5, 
                                     pb210_erosion_loss = 0.5) %>% 
  mutate(core_type = "c. high erosion loss")

a_dates_2 <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 0.01, 
                                     cs137_mobility_factor = 0.01,
                                     cs137_erosion_loss = 0.25, 
                                     pb210_erosion_loss = 0.25) %>% 
  mutate(core_type = "b. mid erosion loss")

a_dates_3 <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 0.01, 
                                     cs137_mobility_factor = 0.01,
                                     cs137_erosion_loss = 0, 
                                     pb210_erosion_loss = 0) %>% 
  mutate(core_type = "a. low erosion loss")


plot1 <- bind_rows(a_dates, a_dates_2, a_dates_3) %>% 
  select(cumVol, supported_pb210_Bq_kg, total_pb210_actual_Bq_kg, unsupported_pb210_actual_Bq_kg, cs137_actual_Bq_kg, core_type) %>% 
  pivot_longer(-c(cumVol, core_type)) %>% 
  mutate(radioisotope = ifelse(grepl("cs137", name), "137Cs", "210Pb"))

ggplot(plot1, aes(x = cumVol, y = value)) +
  geom_point(aes(color = name)) +
  geom_line(aes(color = name)) +
  xlim(c(30,0)) +
  #scale_x_reverse() +
  coord_flip() +
  facet_grid(core_type~radioisotope, scale="free")

# This affects the inventories
inventory_sum <- plot1 %>% 
  filter(name %in% c("cs137_actual_Bq_kg", "unsupported_pb210_actual_Bq_kg")) %>% 
  group_by(core_type, name) %>% 
  summarise(inventory = sum(value)) %>% 
  arrange(name)


(inventory_sum)
```

We can also simulate 

``` {r mobility example}

a_dates_4 <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 5, 
                                     cs137_mobility_factor = 5,
                                     cs137_erosion_loss = 0.25, 
                                     pb210_erosion_loss = 0.25) %>% 
  mutate(core_type = "c. high mobility")

a_dates_5 <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 2.5, 
                                     cs137_mobility_factor = 2.5,
                                     cs137_erosion_loss = 0.25, 
                                     pb210_erosion_loss = 0.25) %>% 
  mutate(core_type = "b. mid mobility")

a_dates_6 <- simulateRadioisotopeDates(core = a_core, cs137_anomaly = 0.6, 
                                     pb210_mobility_factor = 0.01, 
                                     cs137_mobility_factor = 0.01,
                                     cs137_erosion_loss = 0.25, 
                                     pb210_erosion_loss = 0.25) %>% 
  mutate(core_type = "a. low mobility")


plot2 <- bind_rows(a_dates_4, a_dates_5, a_dates_6) %>% 
  select(cumVol, supported_pb210_Bq_kg, total_pb210_actual_Bq_kg, unsupported_pb210_actual_Bq_kg, cs137_actual_Bq_kg, core_type) %>% 
  pivot_longer(-c(cumVol, core_type)) %>% 
  mutate(radioisotope = ifelse(grepl("cs137", name), "137Cs", "210Pb"))

ggplot(plot2, aes(x = cumVol, y = value)) +
  geom_point(aes(color = name)) +
  geom_line(aes(color = name)) +
  xlim(c(30,0)) +
  #scale_x_reverse() +
  coord_flip() +
  facet_grid(core_type~radioisotope, scale="free")

# This affects the inventories
inventory_sum2 <- plot2 %>% 
  filter(name %in% c("cs137_actual_Bq_kg", "unsupported_pb210_actual_Bq_kg")) %>% 
  group_by(core_type, name) %>% 
  summarise(inventory = sum(value)) %>% 
  arrange(name)

(inventory_sum2)

```


